---
output:
  pdf_document: default
  html_document: default
---
# Likelihood
**Author:** Cole Brookson
**Date:** 29 August 2022

<em>Are winter temperatures increasing through time? And if they are, are these warming temperatures responsible for decreasing levels of ice coverage?</em>

One of the first chapters to a book on causal inference that I have recently read is entitled "All you need is regression". I thought this was an accurate way to start this section, because that statement is true! While we cannot cover all the theory of linear regression here, it is genuinely the case that nearly all classical frequentist statistical methods are simply a version of linear regression. A t-test? Actually just linear regression. ANOVA? Yup, also linear regression. Machine learning?? Many types of it are in fact just regression!

So why then, you may ask, did we learn about t-tests before we learned about linear regression in stats class? Well, it turns out that as with almost all quantitative methods discussed in this section, linear regression is frightfully easy to implement, but actually not so simple "under the hood." 

However, getting an intuitive understanding of the basics is very much possible and highly encouraged. Linear regression is probably the most fundamental of all topics you may learn in statistics, and it's worth thinking about it carefully. To be clear, you will likely re-visit this topic many times, likely over the course of years, adding and refreshing pieces of information that will eventually develop into a solid understanding. here, we attempt only to provide a brief introduction to the basics. 

## What is Regression?

A gross simplification of the process of linear regression would be that if we have a series of points between two continuous variables, we want to find a straight line that goes through as many of those points as possible. 

What does it mean to do such a thing? Well, let's start with reminding ourselves, mathematically, what a line is. A straight line is defined by the equation $$ y = mx + b$$ where $b$ is the y-intercept, $m$ is the slope of the line, and $x$ and $y$ are data values. This means that given a point on the x-axis, $x$, a slope $m$, and a y-intercept $b$, we can find the value $y$ that denotes the y-axis value that corresponds to the x-axis value $x$. 

When we talk about linear regression, we'll probably want to discuss this equation in a slightly alterred form: $$y = \beta_0 + \beta_1 x_1$$. We'll discuss what these terms mean in a moment but they relate exactly to a y-intercept ($\beta_0$), a slope ($\beta_1$) and an x-value. 

## Ordinary Least Squares (OLS)

We mentioned above we're putting a line through some points as best we can. But how do we define this "best as we can"? Well, the version of "as best we can" that will be addressed here is the method of least squares. If we imagine a scatter of plots like this: 

```{r, echo = FALSE, warning = FALSE}
x = c(1:100)
y = rep(NA, 100)
for(i in 1:100) {
  y[i] = 1.3 * x[i] + 10 + rnorm(1, mean = 0, sd = 23)
}

df = data.frame(
  x = x, y = y
)

library(tidyverse)
library(ggthemes)
ggplot(df) + 
  geom_point(aes(x = x, y = y), fill = "red", colour = "black", shape = 21, 
             size = 3) + 
  ggthemes::theme_base()
```

It's clear that we can't actually make our line go <em>directly</em> through very many of them, but perhaps we can figure out some way to make the line <em> as close to as many points as possible</em>. We'll do that by imagining we draw a line at random, and calculate the distance between every single point and the line itself. These are called <strong>residuals</strong>. Then we'll square each one, and take the sum of all those squares. 

You might ask, why square it? The short answer to that question is that squaring it makes a large number of mathematical tools available to us. 

It now follows somewhat logically that we want to find a value for the slope of our line, $\beta_1$ that <strong>minimizes</strong> this sum of the squared residuals. What we are doing here is actually a process called <em>estimation</em> wherein we choose some method that allows us to compare candidate values for $\beta_1$ with other values. Ordinary Least Squares is actually just a specific version of Maximum Likelihood Estimation, which we'll discuss below. 

## Linear Model

As in our previous work, we will certainly NOT (!!!!) just jump into writing code, as that's the best way to ensure a poorly done analysis. First, we have our question. Let's frame it as a hypothesis. It seems logical here to state that we think temperature will positively vary with ice cover. So a null hypothesis may be that <em> there is no relationship between air temperature and ice cover </em>. We have decided we want to use linear regression to tackle this challenge, so consider the assumptions of the method: 

1. Linearity - we assume the relationship between our two variables, $X$ and $Y$ is linear

2. Homoskedacity - we assume the variance of teh residual is the same for any value of $X$

3. Normality - We assume that for any fixed value of $X$, the errors of $Y$ normally distributed

4. Randomness - We also assume that samples were collected at random

5. Independence - All errors are independent

If assumptions 1-3 are violated, then we can may be able to use some sort of transformation on our response variable to deal with the problem. 

### Small example 

Let's now get our hands dirty with a bit of an example, a very small one, that we'll go all the way through and then delve into some details in the next example. Very generally, we can think of the process of doing regression as falling into a few steps: 

1) Inspect our data 

2) Perform the regression

3) Assess our assumptions (post-hoc)

4) Inspect our results

5) Interpret & Visualize


Here are the data we will start with: 

```{r}
x = c(1:100)
y = rep(NA, 100)
for(i in 1:100) {
  y[i] = 1.3 * x[i] + 10 + rnorm(1, mean = 0, sd = 23)
}

df = data.frame(
  x = x, y = y
)

head(df)
```

Often it's true when doing regression modelling that it's easier to do our diagnostics of assumptions <em>after</em> we actually fit the model itself as some more simple tools will become available to us. So, let's start with fitting the model. As we mentioned before, R makes it dead simple to do a linear regression in R with the <code>stats::lm()</code> function. 

```{r}
mod <- stats::lm(y ~ x, data = df)
```

That's it! Now let's begin seeing if what our model did was "correct". First, we assumed linearly. We can double check this with a simple scatter plot: 

```{r}
library(tidyverse)
library(ggthemes)
ggplot(df) + 
  geom_point(aes(x = x, y = y), fill = "red", colour = "black", shape = 21, 
             size = 3) + 
  ggthemes::theme_base()
```

This looks sufficiently linear for our purposes. However, we can actually check all of our assumptions visually at once by using the diagnostic plots that are created by simply plotting our model object: 

```{r}
par(mfrow = c(2,2,)) # this ensures all the plots show up at once
plot(mod)
```

So we have four plots here. 

The first plot is the <strong>Residuals vs. Fitted</strong>. This allows us to check the linear relationship assumption. If the red line approximately follows the grey dashed line and there are no distinct patterns, then the variables are linearly related. 

Second is the <strong>Normal QQ</strong>, and we interpret this the same way we interpreted QQ plots previously. If the points follow the line and don't deviate significantly, then the residuals here are normally distributed. 

Third is the <strong>Spread-Location</strong>, which tests the homogeneity of variance in the residuals (our homoskedasticity). Again, we want to see a horizontal red line with "shotgun" (aka randomly spread) points. 

The last is the <strong>Residuals vs. Leverage</strong>. We don't need to look at this plot too carefully right now, but in brief, it's used to identify points having too strong of an influence on the outcome of the analysis. 

Okay so it looks like all our assumptions are satisfied! Let's take a look at our results: 

```{r}
summary(mod)
```

Let's break down each of the components here: 

1. <strong>Call</strong> -- this is the function we used in our model. So to recall, what we're showing here is the that we've regressed <code>y ~ x</code> which we would say as "y regressed against x" or "y regressed with x". 

2. <strong>Residuals</strong> -- These are the residuals. Recall that a <em>residual</em> is the distance from a single point to the regression line at the same x-location. Here's a visual of that: 

```{r}
ggplot(data = df, mapping = aes(x = x, y = y)) + 
  geom_smooth(method = "lm", formula = y ~ x, colour = "red", se = FALSE) +
  ggthemes::theme_base() +
  geom_point(data = df %>% dplyr::filter(
    x %in% c(22, 45, 89)
  ), size = 3, colour = "black", fill = "purple", shape = 21) +
  geom_line(data = data.frame(x = c(22, 22), y = c(
    predict(mod, data.frame(x = 22)), df$y[which(df$x == 22)]
  )), aes(x = x, y = y)) + 
  geom_line(data = data.frame(x = c(45, 45), y = c(
    predict(mod, data.frame(x = 45)), df$y[which(df$x == 45)]
  )), aes(x = x, y = y)) + 
  geom_line(data = data.frame(x = c(89, 89), y = c(
    predict(mod, data.frame(x = 89)), df$y[which(df$x == 89)]
  )), aes(x = x, y = y)) 
```

Each of the lines here going from the points to the the regression line itself are "residuals", or "residual error", the amount of "error" the linear model produces in it's prediction for a given x-value.

Ok back to our <code>summary()</code> call: 

```{r}
summary(mod)
```

the residuals here, by definition, have a mean of zero, and a "perfecty" minimized set of residuals would mean that the median value is zero, which will not really happen, but we want the median to be as close to zero as possible. We also want our minimum and maximum values to be approximately equal in magnitude. 

3) <strong> Coefficients </strong> -- Possibly the part that we'll look at the most, these values are here shown as representing $\beta_0$ and $\beta_1$. So the intercept is the actual y-intecept (remember that $\beta_0$ is the y-intercept), and the $\beta_1$ is the estimate of the coefficient on $x$. We assess whether or not a particular predictor variable is significantly associated with the outcome variable via the "Pr(>|t|)" statement of the p-value. For our purposes, this will be useful, but I encourage you to interpret these p-values with caution, especially when there are multiple variables. 

For our consideration, we take the <em>Estimate</em> as the actual estimated value for the $\beta$s and then the standard errors (SE), which define accuracy of beta coefficients. A larger SE indicates a less certain beta estimate. We then see the t-statistic and the associated p-value, which speak to the statistical significance of the beta coefficients.

We have to recall that for our regression, we are using the $H_0$ that the estimate is zero, which denotes no effect from said coefficient. If any coefficient has a p-value < 0.05, we can infer that the estimate is significantly different from zero. However, this says nothing about the <em>strength</em> of that interaction. 

If we cared to construct a confidence interval around our beta values, we could do so with $\beta +/- 1.96 * SE_{\beta}$, which would give us our 95% CI. 

4) <strong>Model Accuracy</strong> -- We see a number of measures at the bottom, starting with the Residual Standard Error, which contain information about how well our model has fit. 

Starting with the <strong> Residual Standard Error (RSE)</strong>, which essentially represents the average variation of the data points from the fitted regression line. It stands to reason that we want this value to be as small as possible. It comes in handy most when comparing multiple models. Say, for example, we had some other variable $z$ that we think may explain $y$, we could compare the two models $y ~ x$ and $y ~ z$ and see which model has a smaller RSE. This is not the best way to compare models, but it's a good start.

We then see two measures of $R^2$, which is a measure of how well the model fits the data. Specifically, the measure gives the proportion of information (i.e. variation) in the data that can be explained by the model

