---
output:
  pdf_document: default
  html_document: default
---
# Plotting Change Through Time
**Author:** Cole Brookson
**Date:** 20 July 2022


When choosing to plot anything to do with time, the time is almost always displayed on the x-axis. This makes sense as we typically read plots as we read English language text - left to right. Thus, we will always plot time as increasing from left to right on our plots. By FAR the most common way to plot change through time is to plot it as a line plot, optionally with points added in. Less common in EEB-type data (but still useful) are box plots and bar plots.

Let's use an example of a change in temperature and precipitation from Toolik, Alaska, from 1998 to present day to illustrate how to make these plots. 

```{r}
library(tidyverse)
library(lterdatasampler)

df = lterdatasampler::arc_weather
```

## Timeseries Plots (Line Plots)

Line plots, also referred to as timeseries plots colloquially, are plots that have the focal variable of interest on the y-axis, and some time variable on the x-axis. To follow our (LINK TO ITERATIVE PLOTTING) iterative plotting method, we'll start one of these plots and build up.

To start with just the most basic elements (data call, aesthetic mappings, and geom function), we can plot mean air temperature through time: 

```{r}
ggplot() + 
  geom_line(data = df, 
            mapping = aes(x = date, y = mean_airtemp))
```

Okay! Very ugly. Perhaps we can extract some more useful information out of this. But first, let's take care of the easy part and add in the mandatory components (labels, a theme) and then work on getting it more useful. 

```{r}
ggplot() + 
  geom_line(data = df, 
            mapping = aes(x = date, y = mean_airtemp)) 
```

