---
output:
  pdf_document: default
  html_document: default
---
# Renaming
**Author:** Cole Brookson
**Date:** 20 July 2022

```{r, echo=FALSE}
options(tidyverse.quiet = TRUE)
```

It's often the case we want to rename columns in our dataframe. This is achieved very easily. Using an example of a dataset on bison mass, 

```{r}
# load libraries
library(lterdatasampler)
library(tidyverse)

bison <- lterdatasampler::knz_bison 
head(bison)
```

To see what the existing names are, we can call the `names()` function:
```{r}
names(bison)
```

## Rename A Single Column

The TidyR method of renaming is through the `rename()` function from the `dplyr` package. Let's remove the prefix "animal" from "animal_code". In this syntax, the new name comes first, the old name second. 
```{r}
bison <- bison %>% 
  dplyr::rename(code = animal_code)
```
We double check with: 
```{r}
names(bison)
```
And we see it worked. 

## Rename Multiple Columns

We can rename as many columns as we want. 
```{r}
bison <- bison %>% 
  dplyr::rename(
    sex = animal_sex,
    weight = animal_weight,
    yob = animal_yob
  )
names(bison)
```

