---
output:
  pdf_document: default
  html_document: default
---
# Making New Columns
**Author:** Cole Brookson
**Date:** 20 July 2022

```{r, echo=FALSE}
options(tidyverse.quiet = TRUE)
```

While summarizing columns is useful if we want to pare down our dataframe and get a summary, it's often the case we actually just want to add a new column to our dataframe, perhaps based on some set of other columns already in our dataframe. 

In Tidy R, we will almost always use `dplyr::mutate()` for this task. 

## Mutating Columns with Single Values

The simplest use of `mutate()` is to make a new column that is made up of either a pre-existing object or a single value. For examples here, let's turn to a long-term dataset - sugar maple seedlings at Hubbard Brook Experimental Forest. 

```{r}
library(lterdatasampler)
df = lterdatasampler::hbr_maples
head(df)
```

We could for example, want a column in this dataset that tells us what forest these data are from. That would be an easy use of the `mutate()` function as the inclusion of the 